---
title: "Practical Machine Learning - Final Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### Author: Kaustuv Kumar
#### Date : 7th January 2018


## Project Overview

### Background
      
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

      
### Data 

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

## Data Loading and Cleaning

Loading required packages for analysis
```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(randomForest)
```
Download and store the data in a dataframe
```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

Partitioning the training data into two sets

```{r}
inTrain <- createDataPartition(training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]
myTesting <- training[-inTrain, ]
dim(myTraining)
dim(myTesting)
```

Remove non Zero Variance variables

```{r}
nzv <- nearZeroVar(myTraining, saveMetrics=TRUE)
myTraining <- myTraining[,nzv$nzv==FALSE]

nzv<- nearZeroVar(myTesting,saveMetrics=TRUE)
myTesting <- myTesting[,nzv$nzv==FALSE]
# Removing first column
myTraining <- myTraining[c(-1)]
```

Removing the variables having more than 70% NA

```{r}
trainingV3 <- myTraining
for(i in 1:length(myTraining)) {
    if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .7) {
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) == 1)  {
                trainingV3 <- trainingV3[ , -j]
            }   
        } 
    }
}
myTraining <- trainingV3
```

Transform the Validation dataset i.e. MyTesting and actual testing dataset

```{r}
clean1 <- colnames(myTraining)
myTesting <- myTesting[clean1]
dim(myTesting)
clean2 <- colnames(myTraining[, -58]) 
testing <- testing[clean2]
dim(testing)
```

Convert the data into same class

```{r}
for (i in 1:length(testing) ) {
    for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) == 1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}
testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]
```

## Data Modeling with Random Forests

```{r}
set.seed(12345)
modFitB1 <- randomForest(classe ~ ., data=myTraining)
predictionB1 <- predict(modFitB1, myTesting, type = "class")
cmrf <- confusionMatrix(predictionB1, myTesting$classe)
cmrf
plot(modFitB1)
plot(cmrf$table, col = cmrf$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
```



## Data Modeling with Decision Tree

```{r}
set.seed(12345)
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
cmtree <- confusionMatrix(predictionsA1, myTesting$classe)
cmtree
plot(cmtree$table, col = cmtree$byClass, main = paste("Decision Tree Confusion Matrix: Accuracy =", round(cmtree$overall['Accuracy'], 4)))
```

## Prediction on the Test Data

Random Forests gave an Accuracy in the myTesting dataset of 99.75%, which was more accurate that what we got from the Decision Trees.

```{r}
predictionB2 <- predict(modFitB1, testing, type = "class")
predictionB2
```